---
title: "R Day 2"
author: "Dhansree Suraj"
date: "9/18/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
##Reading the data
```{r}
here::here("data", "gapminder5.csv")

gapminder <- read.csv(here::here("data", "gapminder5.csv"))

str(gapminder)
```

## Converting factors to characters

```{r}
gapminder$country <- as.character(gapminder$country)
gapminder$continent <- as.character(gapminder$continent)
str(gapminder)
```

##Creating log variables for for loop
```{r}
obs <- c(1:nrow(gapminder))
for (i in obs) {
  gapminder[i,"log_gdpPercap"] <- log(gapminder[i,"gdpPercap"])
  gapminder[i,"log_pop"] <- log(gapminder[i,"pop"])
}
head(gapminder)
```

##Finding life expectancy for each continent
```{r}
conts <- unique(gapminder$continent)


for (i in conts) {
  lifeExpt <- mean(gapminder$lifeExp[gapminder$continent==i])
  print(paste0(i,": ",lifeExpt))
}

```

##Life Expectancy gap over continents in time
```{r}
years <- unique(gapminder$year)

for(i in conts) {
  print(paste0("Continent: ",i))
  for(j in years) {
    stdLifExp <- sd(gapminder$lifeExp[gapminder$continent==i & gapminder$year==j])
    print(paste0(j,": ",stdLifExp))
  }
}
```

##STD for life expectancy for each year between 1987 and 2002
```{r}
i <- 1987

while(i<=2002){
  sd_le <- sd(gapminder$lifeExp[gapminder$year==i])
  print(paste0(i,": ",sd_le))
  i <- i + 5
}
```


##If else example
```{r}
set.seed(1)
random_year <- sample(years,1)
if(random_year > 1977){
  print(random_year)
} else {
  print("Sorry, random year is less than 1977")
}
```

##Mean population for years greater than or equal to 1987
```{r}
for (i in years){
  if(i<1987){
    print("Sorry, the year is less than 1987")
  } else {
    mean_pop <- mean(gapminder$pop[gapminder$year==i])
    print(paste0(i,": ",mean_pop))
  }
}
```


##Function to mean, median, min, max for life expectancy for a continent
```{r}
get_MeanMedianMinMax <-
  function(df, variable1,variable2,var2val) {
    vals <- df[,variable1][df[,variable2]==var2val]
    mean_val <- mean(vals)
    median_val <- median(vals)
    min_val <- min(vals)
    max_val <- max(vals)
    cat(variable2,": ",var2val,
        "\nMean of ",variable1,": ",mean_val,
        "\nMedian of ",variable1,": ",median_val,
        "\nMin of ",variable1,": ",min_val,
        "\nMax of ",variable1,": ",max_val)
  }
get_MeanMedianMinMax(gapminder,"lifeExp","continent","Asia")
```



##Check if there are any records from 2002 in the dataset
```{r}
if(any(years==2002)){
  print("2002 exists in the data set")
} else {
  print("2002 doesn't exist in the data set")
}
```
##Check if there are any records from 2012 in the dataset
```{r}
if(is.element(2012,years)){
  print("2012 exists in the data set")
} else {
  print("2012 doesn't exist in the data set")
}
```

##Find countries whose mean life expectancy is lower than the overall mean
```{r}
mean_pop <- mean(gapminder$pop)

for(i in unique(gapminder$country)){
  pop_mean_country <- mean(gapminder$pop[gapminder$country==i])
  if(pop_mean_country<mean_pop){
    cou_mean_le <- mean(gapminder$lifeExp[gapminder$country==i])
    print(paste0(i,"'s mean life expectancy is: ",cou_mean_le))
  }
}
```


##Segmenting life expectancy data by continent into <50, 50-70, >70
```{r}
for(i in conts){
  mean_le <- mean(gapminder$lifeExp[gapminder$continent==i])
  if(mean_le<50){
    print(paste0("Mean Life Expectancy for ",i," is less than 50"))
  } else if(mean_le<=70){
    print(paste0("Mean Life Expectancy for ",i," is between 50 and 70"))
  } else {
    print(paste0("Mean Life Expectancy for ",i," is greater than 70"))
  }
}
```


##Function that returns name of columns and class of data in data frame
```{r}
get_ColumnClass <-
  function(df) {
    colname <- NA
    colname <- colnames(df)
    print("Column name: Class")
    i <- colname[1]
    for(i in colname){
      classOfData <- class(df[[i]])
      print(paste0(i,": ",classOfData))
    }
  }

get_ColumnClass(gapminder)
```


##Function to print out mean, sd and optionally median of a vector
```{r}
get_meanSDMed <-
  function(vec, bool=F){
    print(paste0("Mean of vector: ",mean(vec)))
    print(paste0("Standard Deviation of vector: ",sd(vec)))
    if(bool){
      print(paste0("Median of vector: ",median(vec)))
    }
  }

get_meanSDMed(gapminder$pop)
get_meanSDMed(gapminder$pop,T)
```


##Relationship between GDP per capita and Life Expectancy
GDP per capita and life expectancy seem to be highly correlated. As life expectancy increases, so does the GDP per capita.
```{r}
logOfLifeExp <- log(gapminder$lifeExp)
logOfGDP <- log(gapminder$gdpPercap)
fit <- lm(logOfGDP ~ logOfLifeExp)
plot(logOfLifeExp,logOfGDP,
     main="Relationship between the natural logs of life expectancy and GDP per capita",
     xlab="Log of life expectancy",
     ylab="Log of GDP per capita")
lines(logOfLifeExp,predict(fit), col="blue")
```

