---
title: "Final R Exercise"
author: "Dhansree Suraj"
date: "9/19/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Loading data
```{r}
schools <- read.csv(here::here("data","nys_schools.csv"), stringsAsFactors = F)
acs <- read.csv(here::here("data","nys_acs.csv"), stringsAsFactors = F)

```

##Exploring data
```{r}
str(schools)
str(acs)

summary(schools)
summary(acs)
```

##Dealing with -99 values
```{r}
schools[schools==-99 | schools=="-99"] <- NA
library(dplyr)
```

##Creating a categorical variable for poverty in counties
County_per_poverty values below 0.1 (which is the 25th percentile) have been coded as "Low".
County_per_poverty values above 0.15 (which is the 75th percentile) have been coded as "High".
The remaining values have been coded as "Medium".
```{r}
summary(acs$county_per_poverty)
mutate(acs, poverty_category=ifelse(county_per_poverty>=0.15,"High", 
                                     ifelse(county_per_poverty<=0.1,"Low", "Medium")))


acs$poverty_category[is.na(acs$poverty_category)] <- "Medium"
unique(acs$poverty_category)
```

##Scaling scores for math and english
```{r}
schools <- (schools %>% 
  group_by(year) %>% 
  mutate(scaled_ela = scale(mean_ela_score), scaled_math = scale(mean_math_score)))

```

##Merging the datasets
```{r}
acs_schools <- acs %>% 
  merge(schools, by.x=c("county_name","year"), by.y=c("county_name","year"), all.x=T, all.y=T)
```

##Creating Summary tables
```{r}
aggregate(cbind(total_enroll, per_free_lunch, per_reduced_lunch, county_per_poverty) ~ county_name, FUN=mean, na.action = na.omit, data=acs_schools)

acs_schools %>% 
  

#aggregate(cbind(county_per_poverty, per_free_lunch, per_reduced_lunch, scaled_ela, scaled_math) ~ county_name, FUN=mean, na.action = na.omit, data=subset(acs_schools,order(county_per_poverty) %in% head(sort(order(acs_schools$county_per_poverty,na.last = NA)),n=5) | order(county_per_poverty) %in% tail(sort(order(acs_schools$county_per_poverty,na.last = NA)),n=5)))
```

