---
title: "Final R Exercise"
author: "Dhansree Suraj"
date: "9/19/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##1. Loading data
```{r}
schools <- read.csv(here::here("data","nys_schools.csv"), stringsAsFactors = F)
acs <- read.csv(here::here("data","nys_acs.csv"), stringsAsFactors = F)

```

##2. Exploring data
```{r}
str(schools)
str(acs)

summary(schools)
summary(acs)
```

##3. Recoding and variable manipulation
##i. Dealing with -99 values
```{r}
schools[schools==-99 | schools=="-99"] <- NA
library(dplyr)
```

##ii. Creating a categorical variable for poverty in counties
County_per_poverty values below 0.1 (which is the 25th percentile) have been coded as "Low".
County_per_poverty values above 0.15 (which is the 75th percentile) have been coded as "High".
The remaining values have been coded as "Medium".
```{r}
summary(acs$county_per_poverty)
mutate(acs, poverty_category=ifelse(county_per_poverty>=0.15,"High", 
                                     ifelse(county_per_poverty<=0.1,"Low", "Medium")))


acs$poverty_category[is.na(acs$poverty_category)] <- "Medium"
unique(acs$poverty_category)
```

##iii. Scaling scores for math and english
```{r}
schools <- (schools %>% 
  group_by(year) %>% 
  mutate(scaled_ela = scale(mean_ela_score), scaled_math = scale(mean_math_score)))

```

##4. Merging the datasets
```{r}
acs_schools <- acs %>% 
  merge(schools, by.x=c("county_name","year"), by.y=c("county_name","year"), all.x=T, all.y=T)
```

##5. Creating Summary tables
##i. For each county: total enrollment, percent of students qualifying for free or reduced price lunch, and percent of population in poverty.
```{r}
acs_schools %>% 
  select(county_name,county_per_poverty,per_reduced_lunch,per_free_lunch,total_enroll) %>% 
  mutate(reduced_free_lunch = per_free_lunch + per_reduced_lunch) %>% 
  group_by(county_name) %>% 
  summarize(mean_county_per_poverty = mean(county_per_poverty,na.rm=T),
            mean_reduced_free_lunch = mean(reduced_free_lunch,na.rm=T),
            mean_total_enroll = mean(total_enroll,na.rm=T))
```

##ii. Bottom 5 poverty rate counties
```{r}
head(acs_schools %>% 
  select(county_name,county_per_poverty,per_reduced_lunch,per_free_lunch,scaled_ela,scaled_math) %>% 
  mutate(reduced_free_lunch = per_free_lunch + per_reduced_lunch) %>% 
  group_by(county_name) %>% 
  summarize(mean_county_per_poverty = mean(county_per_poverty,na.rm=T),
            mean_reduced_free_lunch = mean(reduced_free_lunch,na.rm=T),
            mean_scaled_ela = mean(scaled_ela,na.rm=T),
            mean_scaled_math = mean(scaled_math,na.rm=T)) %>% 
  arrange(mean_county_per_poverty), n=5)
```

##iii. Top 5 poverty rate counties
```{r}
tail(acs_schools %>% 
  select(county_name,county_per_poverty, per_reduced_lunch, per_free_lunch, scaled_ela,scaled_math) %>% 
  mutate(reduced_free_lunch = per_free_lunch + per_reduced_lunch) %>% 
  group_by(county_name) %>% 
  summarize(mean_county_per_poverty = mean(county_per_poverty,na.rm=T),
            mean_reduced_free_lunch = mean(reduced_free_lunch,na.rm=T),
            mean_scaled_ela = mean(scaled_ela,na.rm=T),
            mean_scaled_math = mean(scaled_math,na.rm=T)) %>% 
  filter(!is.na(county_name)) %>% 
  arrange(mean_county_per_poverty), n=5)
```

##6. Visualizing data
##i. The relationship between access to free/reduced price lunch and test performance, at the school level
```{r}
acs_schools %>% 
  mutate(reduced_free_lunch=per_free_lunch+per_reduced_lunch,scaled_scores=(scaled_ela+scaled_math)/2) %>%
  select(school_cd,school_name, reduced_free_lunch, scaled_scores) %>% 
  group_by(school_cd, school_name) %>%
  summarize (mean_reduced_free_lunch=mean(reduced_free_lunch, na.rm=T), mean_scaled_scores = mean(scaled_scores,na.rm=T)) %>% 
  filter(!is.na(mean_reduced_free_lunch) & !is.na(mean_scaled_scores)) %>% 
  ggplot() +
  geom_point(aes(x=mean_reduced_free_lunch, y=mean_scaled_scores)) +
  labs(title="Relationship between access to free/reduced price lunch and performance score",
       x="% with access to free or reduced price lunch*",
       y="Scaled Math and English scores",
       subtitle = "Calculated at school level",
       caption = "*Calculated by adding % with access to free and % with access to reduced price lunches")

```

##ii. Average test performance across counties with high, low, and medium poverty.
```{r}
acs_schools$poverty_category <- factor(acs_schools$poverty_category, levels = c("Low","Medium","High"))
acs_schools %>% 
  mutate(scaled_scores=(scaled_ela+scaled_math)/2) %>%
  select(county_name,poverty_category,scaled_scores) %>% 
  group_by(county_name, poverty_category) %>%
  summarize (mean_scaled_scores = mean(scaled_scores,na.rm=T)) %>% 
  filter(!is.na(mean_scaled_scores) & !is.na(poverty_category)) %>% 
  ggplot() +
  geom_col(aes(x=county_name, y=mean_scaled_scores, group=poverty_category, fill=poverty_category)) +
  theme_minimal()+
  theme(axis.text.y = element_text(size=6), legend.position="bottom", text = element_text(size=10)) +
  labs(title="Average test performance across counties with high, low and medium poverty",
       x="County Name",
       y="Scaled Math and English scores") + 
  scale_fill_brewer(palette="Greens", name="Poverty Category") +
  coord_flip()
```

